Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands

### 2024-12-13 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd
updates

|              | Variable                  | Source       | Geoprocessing          |
|------------------|--------------------|------------------|------------------|
| Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
|              | Mean Temperature\*        | PRISM        | Basin average          |
|              | Mean Evapotranspiration\* | TerraClimate | Basin average          |
| Geospatial   | Elevation\*               | DEM          | Basin average          |
|              | Area                      | DEM          | Basin average          |
|              | Slope                     | DEM          | Basin average          |
|              | Aspect                    | DEM          | Percent areal coverage |
|              | Land Cover\*              | NLCD         | Percent areal coverage |
|              | Hydrologic Soil Group\*   | SSURGO       | Percent areal coverage |
|              | Geology                   | USGS         | Percent areal coverage |
|              | Karst                     | USGS         | Percent areal coverage |

@@

To check for autocorrelation, we applied the Durbin-Watson test, which revealed significant autocorrelation at four streamgages. Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI. This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998]. Trends with a $\rho \le 0.05$ are considered significant.

@@

![Mean annual BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241203.png){#fig-instrumented-bfi}

@@

#### Trends in BFI {#sec-bfi-trends}

Trends in BFI over the period of record for each streamgage in this analysis are illustrated in @fig-instrumented-trend and @tbl-trends .

@@

| Classification Group               | n   | n_pos | n_neg | perc_pos | perc_neg |
|------------------------------------|-----|-------|-------|----------|----------|
| Precipitation - Monsoon Dominated  | 87  | 8     | 21    | 0.092    | 0.241    |
| Precipitation - Snowmelt Dominated | 118 | 9     | 12    | 0.076    | 0.102    |
| PhysRegion - Basin&Range           | 156 | 14    | 26    | 0.090    | 0.167    |
| PhysRegion - CO Plateau            | 49  | 3     | 7     | 0.061    | 0.143    |
| Climate - Warm-Wet                 | 31  | 2     | 6     | 0.065    | 0.194    |
| Climate - Warm-Dry                 | 55  | 6     | 11    | 0.109    | 0.200    |
| Climate - Cool-Wet                 | 74  | 4     | 9     | 0.054    | 0.122    |
| Climate - Cool-Dry                 | 45  | 5     | 7     | 0.111    | 0.156    |
| Slope - High                       | 102 | 10    | 12    | 0.098    | 0.118    |
| Slope - Low                        | 103 | 7     | 21    | 0.068    | 0.204    |

@@

>>>>

|              | Variable                  | Source       | Geoprocessing          |
|------------------|-------------------|------------------|------------------|
| Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
|              | Mean Temperature\*        | PRISM        | Basin average          |
|              | Mean Evapotranspiration\* | TerraClimate | Basin average          |
| Geospatial   | Elevation\*               | DEM          | Basin average          |
|              | Area                      | DEM          | Basin average          |
|              | Slope                     | DEM          | Basin average          |
|              | Aspect                    | DEM          | Percent areal coverage |
|              | Land Cover\*              | NLCD         | Percent areal coverage |
|              | Hydrologic Soil Group\*   | SSURGO       | Percent areal coverage |
|              | Geology                   | USGS         | Percent areal coverage |
|              | Karst                     | USGS         | Percent areal coverage |

@@

To check for autocorrelation, we applied the Durbin-Watson test, which revealed significant autocorrelation at four streamgages on an annual basis. Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI. This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998]. Trends with a $\rho \le 0.05$ are considered significant.

@@

![Long-term BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241203.png){#fig-instrumented-bfi}

@@

#### Trends in Base Flow & BFI {#sec-bfi-trends}

Trends in BFI over the period of record for each streamgage in this analysis are illustrated in @fig-instrumented-trend and @tbl-trends. Base flow and BFI trends were analyzed across all instrumented sites over their respective periods of record using the Mann–Kendall test. Statistically significant trends were observed in both metrics, with a 72.20% coincidence rate between significant base flow and BFI trends, indicating a strong dependence of BFI on base-flow dynamics.

#### Spatial Trends

TALK ABOUT THE SPATIAL TRENDS IN THE MAP

@@

#### Classification Trends

Classifications presented in @tbl-trends were determined based on precipitation regime, physiographic region, climate, and slope. The dominant precipitation regime (monsoon vs. snowmelt) was identified by analyzing streamflow hydrographs for each station, focusing on peak flow periods during the monsoon season (July–September) and the snowmelt season (March–June). Physiographic region was assigned based on which region the streamgage is in. Climate classifications were defined as warm (above the median temperature of Arizona), cool (below the median temperature), wet (above the median precipitation), and dry (below the median precipitation). Slope was categorized as high (above the median slope) and low (below the median slope).

Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends). While decreasing trends dominate, both increasing and decreasing trends are observed within each classification. Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems may be more vulnerable to declining baseflows. Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to baseflow declines. Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%). This suggests that flatter areas may be more susceptible to baseflow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.

| Classification Group                    | n   | n_pos | n_neg | perc_pos | perc_neg |
|-------------------------|----------|----------|----------|----------|----------|
| Precipitation - Monsoon Dominated       | 87  | 8     | 21    | 0.092    | 0.241    |
| Precipitation - Snowmelt Dominated      | 118 | 9     | 12    | 0.076    | 0.102    |
| Physiographic Region - Basin and Range  | 156 | 14    | 26    | 0.090    | 0.167    |
| Physiographic Region - Colorado Plateau | 49  | 3     | 7     | 0.061    | 0.143    |
| Climate - Warm-Wet                      | 31  | 2     | 6     | 0.065    | 0.194    |
| Climate - Warm-Dry                      | 55  | 6     | 11    | 0.109    | 0.200    |
| Climate - Cool-Wet                      | 74  | 4     | 9     | 0.054    | 0.122    |
| Climate - Cool-Dry                      | 45  | 5     | 7     | 0.111    | 0.156    |
| Slope - High                            | 102 | 10    | 12    | 0.098    | 0.118    |
| Slope - Low                             | 103 | 7     | 21    | 0.068    | 0.204    |

@@

#### Coincident Climate Trends

Given the variations in the period of record across the instrumented network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends in baseflow and BFI in relation to coincident trends in climate variables. Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in baseflow or BFI (@tbl-coincident-trends). This analysis includes both significant and non-significant trends, which is appropriate in environmental studies where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods.

The analysis shows that baseflow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge. Coincidence with AET (actual evapotranspiration) trends (55.12% for baseflow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or baseflow during dry periods. In contrast, temperature trends show lower percentages of coincidence, often opposing baseflow and BFI trends. Specifically, positive (negative) temperature trends are frequently associated with negative (positive) baseflow trends (60.98%) and BFI trends (52.68%). These results highlight the complex interplay between climatic variables and hydrological processes. While precipitation exerts the strongest influence on baseflow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.

|           | **Climate Variable** | **Coincidence Percentage (%)** |
|-----------|:--------------------:|:------------------------------:|
| BFI       |         AET          |             44.88              |
|           |    Precipitation     |             53.17              |
|           |     Temperature      |             47.32              |
| Base Flow |         AET          |             55.12              |
|           |    Precipitation     |             64.88              |
|           |     Temperature      |             39.02              |

: Coincident trends of climate variables (AET, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}
